---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nalanda

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/nalanda)](https://CRAN.R-project.org/package=nalanda)
[![R-CMD-check](https://github.com/centerconflictcooperation/nalanda/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/centerconflictcooperation/nalanda/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/centerconflictcooperation/nalanda/graph/badge.svg)](https://app.codecov.io/gh/centerconflictcooperation/nalanda)
<!-- badges: end -->

## About the Name

The package is named after [Nalanda Mahavihara](https://en.wikipedia.org/wiki/Nalanda), one of the most renowned centers of learning in ancient India. Founded in the 5th century CE, Nalanda was a Buddhist monastic university that attracted scholars from across Asia and became a symbol of knowledge, wisdom, and the pursuit of learning through texts and collaboration.

This name is particularly fitting for a package related to the study of books and prosociality, reflecting the historical significance of Nalanda as a center for both scholarly texts and the cooperative exchange of ideas. The connection resonates with contemporary research on how books and shared learning can foster prosocial behavior and cooperation.

Learn more about related research on books, learning, and prosociality: [Mind and Life Europe - 2024 EVA Recipients & Projects](https://mindandlife-europe.org/2024-eva-recipients-projects/)

## Overview

The **nalanda** package provides tools and utilities for analyzing research data related to books, reading, and prosocial behavior. The package aims to facilitate the exploration of how books and shared learning experiences can foster prosociality and cooperation in communities.

## Installation

You can install the development version of nalanda from [GitHub](https://github.com/) with:
 
``` r
# install.packages("pak")
pak::pak("centerconflictcooperation/nalanda")
```

## Example

The package includes a simple function to explore historical facts about Nalanda University:

```{r example}
library(nalanda)

# Get a random fact about Nalanda University
nalanda()
```
