---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nalanda: R Toolbox to answer the question: Do books really change lives?

## Overview

The **nalanda** package provides tools for simulating, summarizing, and plotting chapter-level AI reading responses. It is designed for workflows that ask whether books shift prosocial attitudes, outgroup warmth, and affective polarization across chapters or across whole books.

## Installation

You can install the development version of nalanda from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
install.packages(
  'nalanda', repos = c(
    'https://centerconflictcooperation.r-universe.dev', 
    'https://cloud.r-project.org'))

```

## Example workflow

The main workflow has three steps:

1. run `run_ai_on_chapters()` to get raw turn-level output
2. convert the raw turns to chapter-level metrics with `compute_run_ai_metrics()`
3. summarize or plot the processed results

This README uses bundled toy data so it renders quickly and does not require live API calls.

```{r, eval=FALSE}
library(nalanda)

raw_turns <- run_ai_on_chapters(
  book_texts = "A short chapter about people from different groups cooperating.",
  groups = c("Democrat", "Republican"),
  context_text = "You are simulating an American adult who politically identifies as a {identity}.",
  question_text = "On a scale from 0 to 100, how warmly do you feel towards {group}s?",
  n_simulations = 2,
  temperature = 0,
  model = "gemini-2.5-flash-lite"
)

chapter_results <- compute_run_ai_metrics(raw_turns)
```

```{r}
library(nalanda)

img_paths <- list(
  Democrat = normalizePath("man/figures/dem.png"),
  Republican = normalizePath("man/figures/rep.png")
)

chapter_results <- compute_run_ai_metrics(toy_run_ai_turns)

head(chapter_results[, c("book", "chapter", "sim", "party", "delta_gap")])
```

Use the processed chapter-level results directly with the time-series plotting helper:

```{r pressure, fig.cap="Synthetic chapter-level trajectories of affective polarization change by party.", fig.width=10, fig.height=7, out.width="100%"}
plot_chapters_over_time(
  chapters = chapter_results,
  dv = "delta_gap",
  group = "party",
  y_label = "Affective Polarization (Delta Gap)",
  plot_subtitle = "Bundled toy data: 2 simulations per party",
  plot_title = TRUE,
  error_bars = FALSE,
  reverse_score = TRUE,
  groups.order = "none",
  facet = "book",
  facets.order = "decreasing",
  line_width = 1.2,
  point_images = img_paths,
  image_size = 0.09
)
```

The same workflow scales to:

- raw turn-level output from `run_ai_on_chapters()` after processing with `compute_run_ai_metrics()`
- summary tables via `summarize_chapter_scores()`
- saved `.rds` simulation outputs
- book-level summaries via `summarize_chapter_scores(..., aggregate_level = "book")`
- visualization with `plot_chapter_trajectories()`, `plot_chapters_over_time()`, and `plot_forest_books()`

For API setup and a live minimal simulation example, see the vignette:

```{r, eval=FALSE}
vignette("getting-started", package = "nalanda")
```

## About the Name

The package is named after [Nalanda Mahavihara](https://en.wikipedia.org/wiki/Nalanda), one of the most renowned centers of learning in ancient India. Founded in the 5th century CE, Nalanda was a Buddhist monastic university that attracted scholars from across Asia and became a symbol of knowledge, wisdom, and the pursuit of learning through texts and collaboration.

This name is particularly fitting for a package related to the study of books and prosociality, reflecting the historical significance of Nalanda as a center for both scholarly texts and the cooperative exchange of ideas. The connection resonates with contemporary research on how books and shared learning can foster prosocial behavior and cooperation.

The package also includes a small helper to explore historical facts about Nalanda University:

```{r example}
library(nalanda)

# Get a random fact about Nalanda University
nalanda()
```

Learn more about related research on books, learning, and prosociality: [Mind and Life Europe - 2024 EVA Recipients & Projects](https://mindandlife-europe.org/2024-eva-recipients-projects/)
