---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nalanda: R Toolbox to answer the question: Do books really change lives?

## Overview

The **nalanda** package provides tools for simulating, summarizing, and plotting chapter-level AI reading responses. It is designed for workflows that ask whether books shift prosocial attitudes, outgroup warmth, and affective polarization across chapters or across whole books.

## Installation

You can install the development version of nalanda from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
install.packages(
  'nalanda', repos = c(
    'https://centerconflictcooperation.r-universe.dev', 
    'https://cloud.r-project.org'))

```

## Example workflow

This example uses a small synthetic dataset so it renders quickly and does not require running live AI simulations.

```{r}
library(nalanda)

summary_chapters <- summarize_chapter_scores(toy_sim_results)

summary_chapters[, c(
  "book", "chapter", "sim",
  "mean_post_outgroup", "mean_delta_gap"
)]
```

Use the summarized output directly with the plotting helpers:

```{r pressure, fig.cap="Synthetic chapter-level summary of post-reading outgroup scores."}
plot_chapter_scores_faceted(
  summary_chapters,
  dv = "post_outgroup",
  y_label = "Mean outgroup rating"
)
```

The same workflow scales to:

- results returned by `run_ai_on_chapters()`
- saved `.rds` simulation outputs
- book-level summaries via `summarize_chapter_scores(..., aggregate_level = "book")`
- visualization with `plot_chapter_trajectories()`, `plot_chapters_over_time()`, and `plot_forest_books()`

For API setup and a live minimal simulation example, see the vignette:

```{r, eval=FALSE}
vignette("getting-started", package = "nalanda")
```

## About the Name

The package is named after [Nalanda Mahavihara](https://en.wikipedia.org/wiki/Nalanda), one of the most renowned centers of learning in ancient India. Founded in the 5th century CE, Nalanda was a Buddhist monastic university that attracted scholars from across Asia and became a symbol of knowledge, wisdom, and the pursuit of learning through texts and collaboration.

This name is particularly fitting for a package related to the study of books and prosociality, reflecting the historical significance of Nalanda as a center for both scholarly texts and the cooperative exchange of ideas. The connection resonates with contemporary research on how books and shared learning can foster prosocial behavior and cooperation.

The package also includes a small helper to explore historical facts about Nalanda University:

```{r example}
library(nalanda)

# Get a random fact about Nalanda University
nalanda()
```

Learn more about related research on books, learning, and prosociality: [Mind and Life Europe - 2024 EVA Recipients & Projects](https://mindandlife-europe.org/2024-eva-recipients-projects/)
