---
title: "Multi-turn survey simulations"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{Multi-turn survey simulations}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nalanda)
```

`run_survey_simulation()` is for simulated respondents who answer a sequence of
survey screens. It differs from `run_prompt_grid()`: prompt-grid variants are
independent conversations, while a survey task is participant × model ×
completion and may preserve conversation state across screens.

```{r, eval=FALSE}
participants <- tibble::tibble(
  participant_id = c("p001", "p002"),
  condition = c("control", "message"),
  age = c(29, 54),
  stimulus = c("A short neutral description.", "A persuasive description.")
)

flow <- tibble::tibble(
  screen_id = c("attention", "outcome"),
  block = c("pre", "post"),
  prompt = c(
    "You are {age} years old. Read: {stimulus}\nWhat phrase stood out?",
    "You are {age} years old in the {condition} condition. Rate your support."
  ),
  response_type = list(
    ellmer::type_object(phrase = ellmer::type_string()),
    ellmer::type_object(support = ellmer::type_number())
  )
)

plan_survey_simulation(participants, flow, model_config = c(flash = "gemini-2.5-flash-lite"))
out <- run_survey_simulation(participants, flow,
  model_config = c(flash = "gemini-2.5-flash-lite"),
  memory = "profile", output_dir = "survey-checkpoints"
)
out$results # one row per answered screen
out$wide    # one row per simulated respondent
```

The participant table is deliberately row-wise. Each row is already an assigned
profile and condition, so it is not expanded into all possible profile-condition
combinations. Any participant column may appear in a prompt as `{column}`.

## Memory is an experimental design choice

With `memory = "conversation"`, a respondent has one continuing chat. This is
convenient for a genuine interview, but it can make answers unusually consistent:
the model can read its earlier answer in a way a human completing separate survey
screens often cannot. `memory = "profile"` starts a fresh chat for each screen,
while still interpolating the participant profile and stimulus into each prompt.
This is usually the closest approximation to independent survey items.

`memory = "block"` preserves context within a named `block` and resets between
blocks. It is useful for a short vignette followed by a separate outcome module.
For each policy, prompts should state the profile/stimulus needed on that screen;
only conversation memory should vary.

Item-per-turn prompting maximizes control over screen wording and per-item
schemas, but costs one call per item. Block prompting reduces calls while keeping
some sections distinct. Whole-survey prompting is cheapest but makes ordering,
missingness, and item-level parsing less transparent. Large simulated survey
benchmarks can therefore be very expensive; always inspect the plan, use
`smoke_n`, and checkpoint before scaling up.

## Branching and resume

An optional `display_if` list-column can contain `TRUE`, `FALSE`, or a function
of `(participant, answers)`. This allows simple deterministic display logic;
the function receives the current participant row and named prior responses.
Random assignment and item randomization should be performed reproducibly while
building `participants` and `survey_flow` and saved with the design. They are not
yet generated by the runner itself, which keeps task identities and resumption
auditable.

Checkpoints are one completed turn at a time. A stopped run resumes without
calling completed screens again, provided the participant, flow, model settings,
completion, and memory policy are unchanged. `on_error = "continue"` records
failed turns for inspection; a later run retries them.
