---
title: "Post-Only Identity Treatments with `simulate_treatment()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Post-Only Identity Treatments with `simulate_treatment()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Purpose

This vignette shows a short workflow for collaborators who currently use
`run_ai_on_chapters()` but only keep the post-treatment responses.

For that use case, `simulate_treatment()` is usually simpler:

1. You define the prompt directly.
2. You can still use `{identity}` and `{intervention_text}` placeholders.
3. You get one row per simulation turn, without pre/post delta columns that you
do not need.

The example below compares `American` and `Brazilian` identities in a
post-only design.

# 1. Set package options

As in other `nalanda` workflows, it is easiest to set the model routing once at
the start of the script.

```{r, eval = TRUE}
library(nalanda)

options(
  nalanda.integration = "vertexai",
  nalanda.base_url = "https://ai-gateway.apps.cloud.rt.nyu.edu/v1/"
)
```

# 2. Define the treatment text and prompt

The intervention text can still be passed separately and inserted into the
prompt with `{intervention_text}`. This is useful when your script generates
many treatment variants programmatically.

```{r, eval = TRUE}
intervention_text <- "Imagine you come across a short community message about 
climate action. It explains that small everyday choices can add up when many 
people participate, and it encourages residents to back clean-energy policies, 
talk with others about practical climate solutions, and choose lower-carbon 
habits when possible. The message ends by saying that collective action now 
can help protect future generations."

prompt <- "{intervention_text}

On a scale from 0 to 100, how accurate do you think this statement is?
Statement: Climate change is a global emergency

On a scale from 0 to 100, how much do you support public policies that reduce greenhouse gas emissions?"
```

Here:

1. `{identity}` will be filled from `groups` via `context_text`.
2. `{intervention_text}` will be filled from the function argument.

# 3. Preview the prompt before running

If you like to inspect prompt text before launching a run, you can build a
single concrete prompt from the same elements you will pass to
`simulate_treatment()`.

```{r, eval = TRUE}
groups <- c("American", "Brazilian")
context_text <- "You are simulating an adult who identifies as {identity}."

prompt_preview <- make_treatment_prompt(
  prompt_template = prompt,
  intervention_text = intervention_text,
  identity_context = gsub("{identity}", groups[1], context_text, fixed = TRUE),
  identity_label = groups[1]
)

cat(prompt_preview)
```

This plays the same role as `make_baseline_prompt()` or `make_post_prompt()`,
but for the one-turn `simulate_treatment()` workflow.

# 4. Run the post-only simulation

```{r, eval = FALSE}
res <- simulate_treatment(
  intervention_text = intervention_text,
  groups = groups,
  context_text = context_text,
  prompt = prompt,
  response_type = ellmer::type_object(
    climate_belief = ellmer::type_number(),
    policy_support = ellmer::type_number()
  ),
  n_simulations = 2,
  temperature = 0,
  model = "gemini-2.5-flash-lite"
)
```

This is the key difference from `run_ai_on_chapters()`:

1. there is no baseline turn,
2. there are no `pre_*` or `delta_*` columns,
3. the output matches the post-only design directly.

If you want a multi-turn design later, `simulate_treatment()` can also take a
prompt vector of length greater than 1, with one element per turn.

# 5. Inspect the raw output

Each row is one simulated response for one identity and one simulation draw.

```{r, echo = FALSE}
example_res <- tibble::tibble(
  chapter = "intervention_1",
  sim = c(1L, 2L, 1L, 2L),
  identity = c("American", "American", "Brazilian", "Brazilian"),
  turn_index = 1L,
  turn_type = "turn_1",
  climate_belief = c(84, 82, 76, 74),
  policy_support = c(71, 69, 62, 60)
)

knitr::kable(example_res)
```

# 6. Summarize by identity

Use `summarize_treatment_results()` to compute mean scores by identity.

```{r, eval = FALSE}
summary_by_identity <- summarize_treatment_results(
  res,
  by_identity = TRUE
)

summary_by_identity
```

```{r, echo = FALSE}
example_summary <- tibble::tibble(
  ch = "intervention_1",
  id = c("American", "Brazilian"),
  turn = "turn_1",
  n = c(2L, 2L),
  m_climate_belief = c(83.0, 75.0),
  sd_climate_belief = c(1.4, 1.4),
  m_policy_support = c(70.0, 61.0),
  sd_policy_support = c(1.4, 1.4),
  idx = 1L
)

knitr::kable(example_summary)
```

# 7. Compare another group to Americans

If `American` is your reference group, a simple downstream comparison is:

```{r, eval = FALSE}
library(dplyr)
library(tidyr)

comparison_vs_american <- summary_by_identity |>
  select(
    identity,
    mean_climate_belief,
    mean_policy_support
  ) |>
  pivot_wider(
    names_from = identity,
    values_from = c(mean_climate_belief, mean_policy_support)
  ) |>
  mutate(
    climate_belief_gap_vs_american =
      mean_climate_belief_Brazilian - mean_climate_belief_American,
    policy_support_gap_vs_american =
      mean_policy_support_Brazilian - mean_policy_support_American
  )

comparison_vs_american
```

```{r, echo = FALSE}
comparison_vs_american <- tibble::tibble(
  climate_us = 83.0,
  climate_br = 75.0,
  policy_us = 70.0,
  policy_br = 61.0,
  gap_climate = -8.0,
  gap_policy = -9.0
)

knitr::kable(comparison_vs_american)
```

# When to use which function

Use `run_ai_on_chapters()` when you want a true pre/post design and plan to use
within-agent change scores such as `delta_outgroup` or `delta_gap`.

Use `simulate_treatment()` when:

1. the design is post-only,
2. the treatment is not really a book chapter,
3. you want direct control over the prompt wording, and
4. you want to insert `{intervention_text}` programmatically.
